---
author: "Assignment A8 by Shreysa Sharma"

output:
  pdf_document
---

```{r setup, echo=FALSE, results='hide',message=FALSE, warning=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath(getwd()))
```

##Program Design

The goal of this assignment was to perform provide visualization mean delay of the five most active airlines and for the five most active airports in the country. I have used `1 Map-Reduce job` that does the initial cleaning of the airline dateset and then gets the normalized delay(with respect to the scheduled length of each flight) and mean delay for each month for a particular year. The output is a results.csv file placed in the `output` folder that has airportId and year or airlineId and year as key and value as NormalizedDelay for the year and for each month seperated by `","` starting from Jan.
##### Plot 1: Most Active Airlines in the period 2011-2015 with their Mean Delay
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=10}
library(plyr)
library(knitr)
library(sqldf)
library(ggplot2)

# AIRLINE
airline_files = read.csv("./output/results.csv", header = FALSE)
#airline_files = dir(path="./output_small/results.csv", full.names=TRUE)
colnames(airline_files) <- c("airlineId", "year", "month","type", "NormalizedDelay","count")
airline_data <- airline_files[airline_files$type == '0',]
airport_data <-airline_files[airline_files$type == '1',]
write.csv(file="./output/airlineData.csv", x=airline_data)
write.csv(file="./output/airportData.csv", x=airport_data)
airline_id <- read.csv("L_AIRLINE_ID.csv")
airport_id <- read.csv("L_AIRPORT_ID.csv")

grp_by_airline <- aggregate(airline_data$count,by=list(airlineId=airline_data$airlineId), FUN=sum)
dfMeanSorted<-arrange(grp_by_airline, desc(x))
TopAirline=dfMeanSorted[1:5,]
grp_by_airport <- aggregate(airport_data$count,by=list(airlineId=airport_data$airlineId), FUN=sum)
dfMeanSortedAirport<-arrange(grp_by_airport, desc(x))
TopAirport=dfMeanSortedAirport[1:5,]


active_airlines_table <- sqldf("select F.airlineId, F.year,F.month, F.NormalizedDelay,T.x as count from airline_data F JOIN TopAirline T where F.airlineId = T.airlineId")

active_airlines <- sqldf("select N.Description, F.airlineId, F.year, F.month, F.NormalizedDelay, F.count from active_airlines_table F JOIN airline_id N where N.Code = F.airlineId")

active_airports_table <- sqldf("select F.airlineId, F.year,F.month, F.NormalizedDelay,T.x as count from airport_data F JOIN TopAirport T where F.airlineId = T.airlineId")

active_airports <- sqldf("select N.Description, F.airlineId, F.year, F.month, F.NormalizedDelay, F.count from active_airports_table F JOIN airport_id N where N.Code = F.airlineId")
#active_airlines_2015 <- sqldf("select * from active_airlines where year=2015 or year =2014 or year=2013 or year=2012 or year= 2011")
active_airlines_715 <- sqldf("select * from active_airlines where year in(1987, 1988, 1989, 1990, 1991, 1992, 1993, 1994, 1995, 1996, 1997, 1998, 1999, 2000,2001, 2002, 2003,2004, 2005, 2006, 2007, 2008, 2009, 2010,2011, 2012, 2013, 2014, 2015)")
#active_airlines_715 <- sqldf("select * from active_airlines where year in(1987, 1988, 1989, 1990, 1991, 1992, 1993, 1994, 1995, 1996, 1997, 1998)")
active_airlines_806 <- sqldf("select * from active_airlines where year in(1998, 1999, 2000, 2001,2002, 2003, 2004, 2005, 2006)")
active_airlines_8997 <- sqldf("select * from active_airlines where year in(1989, 1990, 1991, 1992,1993, 1994, 1995, 1996, 1997)")
active_airports_715 <- sqldf("select * from active_airports where year in(2007, 2008, 2009, 2010,2011, 2012, 2013, 2014, 2015)")
active_airports_806 <- sqldf("select * from active_airports where year in(1998, 1999, 2000, 2001,2002, 2003, 2004, 2005, 2006)")
active_airports_8997 <- sqldf("select * from active_airports where year in(1989, 1990, 1991, 1992,1993, 1994, 1995, 1996, 1997)")



grp_by_airline <- aggregate(airline_data$count,by=list(airlineId=airline_data$airlineId), FUN=sum)
dfMeanSorted<-arrange(grp_by_airline, desc(x))
TopAirlineAfter=dfMeanSorted[1:6,]
active_airlines_tableAfter <- sqldf("select F.airlineId, F.year,F.month, F.NormalizedDelay,T.x as count from airline_data F JOIN TopAirlineAfter T where F.airlineId = T.airlineId and year>2009")
active_airlinesAfter <- sqldf("select N.Description, F.airlineId, F.year, F.month, F.NormalizedDelay, F.count from active_airlines_tableAfter F JOIN airline_id N where N.Code = F.airlineId")


active_airports_1988 <- sqldf("select * from active_airports where year =1992")
active_airports_1989 <- sqldf("select * from active_airports where year =1993")
ggplot(active_airports_1988,
             aes(x=month, y=NormalizedDelay)) + geom_bar(aes(fill=factor(Description), colour=factor(Description)), position = "dodge", stat="identity")+facet_wrap(~ year)+ scale_x_discrete(limits=1:12,breaks=seq(0,12,3)) + theme(legend.position="top")


ggplot(active_airports_1989,
             aes(x=month, y=NormalizedDelay)) + geom_bar(aes(fill=factor(Description), colour=factor(Description)), position = "dodge", stat="identity")+facet_wrap(~ year)+ scale_x_discrete(limits=1:12,breaks=seq(0,12,3)) + theme(legend.position="top")

active_airports_1990 <- sqldf("select * from active_airports where year =1994")
active_airports_1991 <- sqldf("select * from active_airports where year =1991")
ggplot(active_airports_1990,
             aes(x=month, y=NormalizedDelay)) + geom_bar(aes(fill=factor(Description), colour=factor(Description)), position = "dodge", stat="identity")+facet_wrap(~ year)+ scale_x_discrete(limits=1:12,breaks=seq(0,12,3)) + theme(legend.position="top")


ggplot(active_airports_1991,
             aes(x=month, y=NormalizedDelay)) + geom_bar(aes(fill=factor(Description), colour=factor(Description)), position = "dodge", stat="identity")+facet_wrap(~ year)+ scale_x_discrete(limits=1:12,breaks=seq(0,12,3)) + theme(legend.position="top")

#ggplot(active_airlines_715,
#             aes(x=month, y=NormalizedDelay)) + geom_point(aes(fill=factor(Description), colour=factor(Description)), position = "dodge", stat="identity")+facet_wrap(Description ~ year)+ scale_x_discrete(limits=1:12,breaks=seq(0,12,3)) + theme(legend.position="top")

#ggplot(active_airlines_8997, aes(x=month, y=NormalizedDelay)) + 
#  geom_point(aes(fill=factor(Description), colour=factor(Description)), position = "dodge", stat="identity") +
#  facet_wrap(~year) + 
#  scale_x_discrete(limits=1:12,breaks=seq(0,12,3)) +
#  theme(legend.position="top") #+
  #coord_fixed(ratio = 0.05)
```


```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=11}
#ggplot(active_airlines_806,
#             aes(x=month, y=NormalizedDelay)) + geom_point(aes(fill=factor(airlineId), colour=factor(airlineId)), position = "dodge", #stat="identity")+facet_wrap(~year)+ scale_x_discrete(limits=1:12,breaks=seq(0,12,3))  + theme(legend.position="top")

#ggplot(active_airlines_715,
#             aes(x=month, y=NormalizedDelay)) + geom_point(aes(fill=factor(airlineId), colour=factor(airlineId)), position = "dodge", stat="identity")+facet_wrap(~year)+ scale_x_discrete(limits=1:12,breaks=seq(0,12,3)) + theme(legend.position="top")

#ggplot(active_airports_8997,
#             aes(x=month, y=NormalizedDelay)) + geom_point(aes(fill=factor(airlineId), colour=factor(airlineId)), position = "dodge", #stat="identity")+facet_wrap(~year)+ scale_x_discrete(limits=1:12,breaks=seq(0,12,3)) + theme(legend.position="top")
#ggplot(active_airports_806,
#             aes(x=month, y=NormalizedDelay)) + geom_point(aes(fill=factor(airlineId), colour=factor(airlineId)), position = "dodge", stat="identity")+facet_wrap(~year)+ scale_x_discrete(limits=1:12,breaks=seq(0,12,3)) + theme(legend.position="top")
#ggplot(active_airports_715,
 #            aes(x=month, y=NormalizedDelay)) + geom_point(aes(fill=factor(airlineId), colour=factor(airlineId)), position = "dodge", #stat="identity")+facet_wrap(~year)+ scale_x_discrete(limits=1:12,breaks=seq(0,12,3)) + theme(legend.position="top")

```


```{r}



grp_by_airline <- aggregate(airline_data$count,by=list(airlineId=airline_data$airlineId), FUN=sum)
dfMeanSortedAfter<-arrange(grp_by_airline, desc(x))
TopAirlineAfter=dfMeanSortedAfter[1:6,]
active_airlines_tableAfter <- sqldf("select F.airlineId, F.year,F.month, F.NormalizedDelay,T.x as count from airline_data F JOIN TopAirlineAfter T where F.airlineId = T.airlineId and year>2009")
active_airlinesAfter <- sqldf("select N.Description, F.airlineId, F.year, F.month, F.NormalizedDelay, F.count from active_airlines_tableAfter F JOIN airline_id N where N.Code = F.airlineId")




```
The Plots 1-? represent the Normalized delay for the Top 5 airlines. The available data is from Dec 1987 to Oct 2015. For the period 1987 to 1990 US Airways topped the list for delays, however towards the end of 1990 Southwest and Delta were more delayed. Thereafter, for a majority period from 1991 to 1998 Delta was the most delayed flight, closely follwed by Southwest and US airways. It is interesting to note that **Northwest Airlines** was the least delayed among these 5 top airlines from 1987 to 2004 but after 2004 it topped the list for delays till 2009. It can be seen that after 2009 Northwest Airlines does not appear in the graph as it was merged with Delta airlines in 2010 Jan due to bankcruptcy and closed down operations as Northwest in 2009 Dec.

After 1998 Southwest has topped the list for delays among airlines while American Airlines which appeared in the top 5 busiest airlines after NorthWest stopped its services has the least delay among the set.

It was also noticed that after 2011 the mean delays for all the airlines have reduced considerably.


The Plots 1-? represent the Normalized delay for the Top 5 airports. The top 5 busiest airports are: **Hartsfield-Jackson Atlanta International, Chicago O'Hare International, Los Angeles, CA: Los Angeles International, Phoenix Sky Harbor International and Denver International Airport**. The available data is from Dec 1987 to Oct 2015. For the period 1987 to 1996 `Hartsfield-Jackson Atlanta International` Airport was the airport with most delays however after 1996 `Chicago O'Hare International` Airport has been the most prominent out of the top 5 busiest airports. `Denver International Airport` was the one with least delays among the top 5 busiest airports till 2010 however post that `Phoenix Sky Harbor International Airport` has the minimum delays out of the set.  


#### Conclusion

The exercise helped in observing few interesting facts. 

Delta Airlines was the most delayed airline from 1991 to 1998 but post that Southwest airline has been the most delayed airline among the top 5 busiest airlines while NorthWest was the one with minimum delays in the set till 2004 and after it merged with delta, American Airlines joined the group of top 5 busiest airlines and had the least delay.

NorthWest Airline was one of the top 5 busiest airline and also had minimum delays among the this set but close to the year when it was shut down the mean delay of the airline grew with it being the one with most delays in the years 2005-2009.

The mean delays for the airlines were considerably reduced after 2011.

It was noticed that the maximum delays for all airlines and airports were in the month of `December` which could primarily be because of December being the holiday month, most people travel during that time. Additionally, `September` has the least number of delays which could be because schools reopen after summer vacations and less people travel then. 